Note on Linear Biases in MORUS

### 2018-04-11 morus.tex
typo fix

Our main result is the existence of a linear trail linking ciphertext bits spanning five consecutive encryption steps. As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps. Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design. Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initalize the inner state of MORUS are irrelevant: the bias is independent of their value.

@@

Of course, the overall trail resulting from all of the previous combinations is quite complex, especially since $\gamma^t_i$, $\delta^t_i$, and $\varepsilon^t_i$ each require two copies of the preceding trail fragment in consecutive steps: that is, $\varepsilon^t_i$ requires two approximations of $S_2$, which requires using $\delta^t_i$ twice; and $\delta^t_i$ in turn requires using $\gamma^t_i$ twice, which itself requires using $\alpha^t_i$ and $\beta^t_i$ twice. Then $\alpha^t_i$ is used one final time to close the trail. One may naturally wonder if some components of this trail are in conflict. In particular, products of bits from registers $S_2$ and $S_3$ are approximated multiple times, by $\alpha^t_i$, $\beta^t_i$ and $\gamma^t_i$. To address this concern, and ensure that all approximations along the trail are in fact compatible, we now compute the full trail equation explicitely.

@@

The equation corresponding to each of the five trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, $\varepsilon^t_i$ may be written explicitely as $A^t_i$, $B^t_i$, $C^t_i$, $D^t_i$, $E^t_i$ as follows. For each equation, we write on the left-hand side of the equality the biased linear combination used in the trail; and on the right-hand side, the remainder of the equation, which must have non-zero bias (in all cases the bias is $2^{-1}$).
\begin{alignat*}{2}
A^t_i:\quad&& C^t_i \oplus S^{t+1}_{0,i+b_0} &= S^t_{1,i} \oplus S^t_{3,i} \oplus S^t_{1,i} \cdot S^t_{2,i} \oplus S^t_{2,i} \cdot S^t_{3,i}\\
B^t_i:\quad&& C^t_i \oplus S^t_{0,i} \oplus S^t_{1,i} &= S^t_{2,i} \cdot S^t_{3,i} \\
C^t_i:\quad&& S^t_{1,i} \oplus S^{t+1}_{1,i+b_1} \oplus S^t_{4,i} &= S^t_{2,i} \cdot S^t_{3,i}\\
D^t_i:\quad&& S^t_{4,i} \oplus S^{t+1}_{4,i+b_4} \oplus S^{t+1}_{2,i} &= S^{t+1}_{0,i} \cdot S^{t+1}_{1,i}\\
E^t_i:\quad&& S^t_{2,i} \oplus S^{t+1}_{2,i+b_2} \oplus S^{t+1}_{0,i} &= S^t_{3,i} \cdot S^t_{4,i}
\end{alignat*}

@@

That the bias is so high may be suprising: indeed, the full trail uses trail fragment $\varepsilon^t_i$ 1 time, $\delta^t_i$ 2 times, $\gamma^t_i$ 4 times, $\beta^t_i$ 8 times, and $\alpha^t_i$ 9 times. Since each trail fragment has a weight of 1, this would suggest that the total weight should be $1+2+4+8+9 = 24$ rather than 16. However, when combining trail fragments $\beta_i$ and $\gamma_i$, notice that the same AND is computed at the same step between registers $S_2$ and $S_3$ (equivalently, notice that the right-hand side of equations $B^t_i$ and $C^t_i$ is equal). In both cases it is approximated by zero. When XORing the corresponding equations, these two ANDs cancel each other, which saves two AND gates. Since $\gamma^t_i$ is used four times in the course of the full trail, this results in saving 8 AND gates overall, which explains why the final bias is $2^{-16}$ rather than $2^{-24}$.

@@

Thus, translating the trail from one of the \cipher{MiniMORUS} ciphers to the corresponding full \cipher{MORUS} cipher amounts to making every linear combination symmetric---indeed, that was the point of introducing \cipher{MiniMORUS} in the first place.
Concretely, in order to build the full trail equation for \cipher{MORUS}, we write symmetric versions of equations $A^t_i$, $B^t_i$, $C^t_i$, $D^t_i$, $E^t_i$ from \Cref{sec:minitraileq}, and then combine them in exactly the same way as before.
In this way, the biased linear combination on \cipher{MiniMORUS640} given in \Cref{sec:minitraileq}, namely:
\begin{align*}
&C^0_{27} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^2_{1} \oplus C^2_{13} \oplus C^2_{15}\\
\oplus\; &C^2_{27} \oplus C^2_{31} \oplus C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^4_{19}
\end{align*}
utlimately yields the following biased symmetrized linear combination on the full \cipher{MORUS640}:
\begin{align*}
&C^0_{27} \oplus C^0_{59} \oplus C^0_{91} \oplus C^0_{123} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^1_{32} \oplus C^1_{34}\\
\oplus\; & C^1_{40} \oplus C^1_{58} \oplus C^1_{64} \oplus C^1_{66} \oplus C^1_{72} \oplus C^1_{90} \oplus C^1_{96} \oplus C^1_{98} \oplus C^1_{104} \oplus C^1_{122}\\
\oplus\; & C^2_{1} \oplus C^2_{13} \oplus C^2_{15} \oplus C^2_{27} \oplus C^2_{31} \oplus C^2_{33} \oplus C^2_{45} \oplus C^2_{47} \oplus C^2_{59} \oplus C^2_{63}\\
\oplus\; & C^2_{65} \oplus C^2_{77} \oplus C^2_{79} \oplus C^2_{91} \oplus C^2_{95} \oplus C^2_{97} \oplus C^2_{109} \oplus C^2_{111} \oplus C^2_{123} \oplus C^2_{127}\\
\oplus\; & C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^3_{38} \oplus C^3_{44} \oplus C^3_{46} \oplus C^3_{52} \oplus C^3_{70} \oplus C^3_{76}\\
\oplus\; & C^3_{78} \oplus C^3_{84} \oplus C^3_{102} \oplus C^3_{108} \oplus C^3_{110} \oplus C^3_{116} \oplus C^4_{19} \oplus C^4_{51} \oplus C^4_{83} \oplus C^4_{115}.
\end{align*}
We refer the reader to \Cref{sec:traileq} for the corresponding linear combination on \cipher{MORUS1280}.

@@

In the actual trail for (either version of) full \cipher{MORUS}, this exact situation occurs when combining trail fragments $\beta^t_i$ and $\gamma^t_i$. Indeed, $\beta^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i}$, while $\gamma^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i-w}$ (cf. \Cref{fig:trailcollision}). While in \cipher{MiniMORUS}, these terms cancel out, in the full \cipher{MORUS}, when adding all symmetric copies of the trail, we end up with the sum:
\begin{align}
&S^t_{2,i} \cdot S^t_{3,i} \oplus S^t_{3,i} \cdot S^t_{2,i+w}
\oplus S^t_{2,i+w} \cdot S^t_{3,i+w} \oplus S^t_{3,i+w} \cdot S^t_{2,i+2w}\notag\\
\oplus\; &S^t_{2,i+2w} \cdot S^t_{3,i+2w} \oplus S^t_{3,i+2w} \cdot S^t_{2,i+3w}
\oplus S^t_{2,i+3w} \cdot S^t_{3,i+3w} \oplus S^t_{3,i+3w} \cdot S^t_{2,i}.\label{eq:8circle}
\end{align}
It may be observed that the products occuring in the equation above involve 8 terms in a ring. The weight of this expression can be computed by brute force, and is equal to $3$.

@@

In the end, for \cipher{MORUS1280}, since the trail fragment $\gamma^t_i$ is used four times, the phenomenon we just described adds a contribution of $4 \cdot 3 = 12$ to the overall weight of the full trail. This results in a total weight of $4 \cdot 16 + 12 = 76$ (recall that the weight of the trail on \cipher{MiniMORUS1280} is 16). We have confirmed this bias by explicitely computing the full trail equation in \Cref{sec:traileq}, and evaluating its exact weight like we did for \cipher{MiniMORUS} in \Cref{sec:minibias}. That is, since the equation is quadratic, we may view it as a graph, which we split into connected components; we then compute the bias of each connected component separately by brute force, and then add up the weights of all components per the piling-up lemma. In the end, the full trail equation given in \Cref{sec:traileq} yields a weight of 76 for the full trail on \cipher{MORUS1280}.

In the case of \cipher{MORUS640}, collisions between rotation constants further complicate the analysis. Specifically, when using trail fragment $\beta^t_i$, the term $S^t_{2,i} \cdot S^t_{3,i}$ occurs. As explained previously, a partial collision with the term $S^t_{2,i} \cdot S^t_{3,i-w}$ from trail fragment $\gamma^t_i$ results in \Cref{eq:8circle}. However trail fragment $\alpha^t_{i+d}$ is once used in the course of the full trail with an offest of $d = b_1+b_4-b_0-b_2$ (relative to $\gamma^t_i$), which in the case \cipher{MORUS640} is equal to $31+13-5-7 = 0 \;\text{mod}\; 32$. This creates another term $S^t_{2,i} \cdot S^t_{3,i}$, which ultimately destroys one of the four occurences of \Cref{eq:8circle}. In the end, when computing the full trail equation on \cipher{MORUS640}, we get that the weight of the trail is 73 (cf. \Cref{sec:traileq}).

>>>>

Our main result is the existence of a linear trail linking ciphertext bits spanning five consecutive encryption steps. As a consequence of this trail, we exhibit a biased linear combination of ciphertext bits, which holds across any given sequence of five consecutive encryption steps. Technically, this assumes that the inner state of the cipher at the start of the trail is uniformly random, but this is a fair assumption since the inner state should be indistinguishable from uniform randomness by design. Moreover, that is the only assumption required; in particular, the secret key or the nonce used to initialize the inner state of MORUS are irrelevant: the bias is independent of their value.

@@

Of course, the overall trail resulting from all of the previous combinations is quite complex, especially since $\gamma^t_i$, $\delta^t_i$, and $\varepsilon^t_i$ each require two copies of the preceding trail fragment in consecutive steps: that is, $\varepsilon^t_i$ requires two approximations of $S_2$, which requires using $\delta^t_i$ twice; and $\delta^t_i$ in turn requires using $\gamma^t_i$ twice, which itself requires using $\alpha^t_i$ and $\beta^t_i$ twice. Then $\alpha^t_i$ is used one final time to close the trail. One may naturally wonder if some components of this trail are in conflict. In particular, products of bits from registers $S_2$ and $S_3$ are approximated multiple times, by $\alpha^t_i$, $\beta^t_i$ and $\gamma^t_i$. To address this concern, and ensure that all approximations along the trail are in fact compatible, we now compute the full trail equation explicitly.

@@

The equation corresponding to each of the five trail fragments $\alpha^t_i$, $\beta^t_i$, $\gamma^t_i$, $\delta^t_i$, $\varepsilon^t_i$ may be written explicitly as $A^t_i$, $B^t_i$, $C^t_i$, $D^t_i$, $E^t_i$ as follows. For each equation, we write on the left-hand side of the equality the biased linear combination used in the trail; and on the right-hand side, the remainder of the equation, which must have non-zero bias (in all cases the bias is $2^{-1}$).
\begin{alignat*}{2}
A^t_i:\quad&& C^t_i \oplus S^{t+1}_{0,i+b_0} &= S^t_{1,i} \oplus S^t_{3,i} \oplus S^t_{1,i} \cdot S^t_{2,i} \oplus S^t_{2,i} \cdot S^t_{3,i}\\
B^t_i:\quad&& C^t_i \oplus S^t_{0,i} \oplus S^t_{1,i} &= S^t_{2,i} \cdot S^t_{3,i} \\
C^t_i:\quad&& S^t_{1,i} \oplus S^{t+1}_{1,i+b_1} \oplus S^t_{4,i} &= S^t_{2,i} \cdot S^t_{3,i}\\
D^t_i:\quad&& S^t_{4,i} \oplus S^{t+1}_{4,i+b_4} \oplus S^{t+1}_{2,i} &= S^{t+1}_{0,i} \cdot S^{t+1}_{1,i}\\
E^t_i:\quad&& S^t_{2,i} \oplus S^{t+1}_{2,i+b_2} \oplus S^{t+1}_{0,i} &= S^t_{3,i} \cdot S^t_{4,i}
\end{alignat*}

@@

That the bias is so high may be surprising: indeed, the full trail uses trail fragment $\varepsilon^t_i$ 1 time, $\delta^t_i$ 2 times, $\gamma^t_i$ 4 times, $\beta^t_i$ 8 times, and $\alpha^t_i$ 9 times. Since each trail fragment has a weight of 1, this would suggest that the total weight should be $1+2+4+8+9 = 24$ rather than 16. However, when combining trail fragments $\beta_i$ and $\gamma_i$, notice that the same AND is computed at the same step between registers $S_2$ and $S_3$ (equivalently, notice that the right-hand side of equations $B^t_i$ and $C^t_i$ is equal). In both cases it is approximated by zero. When XORing the corresponding equations, these two ANDs cancel each other, which saves two AND gates. Since $\gamma^t_i$ is used four times in the course of the full trail, this results in saving 8 AND gates overall, which explains why the final bias is $2^{-16}$ rather than $2^{-24}$.

@@

Thus, translating the trail from one of the \cipher{MiniMORUS} ciphers to the corresponding full \cipher{MORUS} cipher amounts to making every linear combination symmetric---indeed, that was the point of introducing \cipher{MiniMORUS} in the first place.
Concretely, in order to build the full trail equation for \cipher{MORUS}, we write symmetric versions of equations $A^t_i$, $B^t_i$, $C^t_i$, $D^t_i$, $E^t_i$ from \Cref{sec:minitraileq}, and then combine them in exactly the same way as before.
In this way, the biased linear combination on \cipher{MiniMORUS640} given in \Cref{sec:minitraileq}, namely:
\begin{align*}
&C^0_{27} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^2_{1} \oplus C^2_{13} \oplus C^2_{15}\\
\oplus\; &C^2_{27} \oplus C^2_{31} \oplus C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^4_{19}
\end{align*}
ultimately yields the following biased symmetrized linear combination on the full \cipher{MORUS640}:
\begin{align*}
&C^0_{27} \oplus C^0_{59} \oplus C^0_{91} \oplus C^0_{123} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^1_{32} \oplus C^1_{34}\\
\oplus\; & C^1_{40} \oplus C^1_{58} \oplus C^1_{64} \oplus C^1_{66} \oplus C^1_{72} \oplus C^1_{90} \oplus C^1_{96} \oplus C^1_{98} \oplus C^1_{104} \oplus C^1_{122}\\
\oplus\; & C^2_{1} \oplus C^2_{13} \oplus C^2_{15} \oplus C^2_{27} \oplus C^2_{31} \oplus C^2_{33} \oplus C^2_{45} \oplus C^2_{47} \oplus C^2_{59} \oplus C^2_{63}\\
\oplus\; & C^2_{65} \oplus C^2_{77} \oplus C^2_{79} \oplus C^2_{91} \oplus C^2_{95} \oplus C^2_{97} \oplus C^2_{109} \oplus C^2_{111} \oplus C^2_{123} \oplus C^2_{127}\\
\oplus\; & C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^3_{38} \oplus C^3_{44} \oplus C^3_{46} \oplus C^3_{52} \oplus C^3_{70} \oplus C^3_{76}\\
\oplus\; & C^3_{78} \oplus C^3_{84} \oplus C^3_{102} \oplus C^3_{108} \oplus C^3_{110} \oplus C^3_{116} \oplus C^4_{19} \oplus C^4_{51} \oplus C^4_{83} \oplus C^4_{115}.
\end{align*}
We refer the reader to \Cref{sec:traileq} for the corresponding linear combination on \cipher{MORUS1280}.

@@

In the actual trail for (either version of) full \cipher{MORUS}, this exact situation occurs when combining trail fragments $\beta^t_i$ and $\gamma^t_i$. Indeed, $\beta^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i}$, while $\gamma^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t_{3,i-w}$ (cf. \Cref{fig:trailcollision}). While in \cipher{MiniMORUS}, these terms cancel out, in the full \cipher{MORUS}, when adding all symmetric copies of the trail, we end up with the sum:
\begin{align}
&S^t_{2,i} \cdot S^t_{3,i} \oplus S^t_{3,i} \cdot S^t_{2,i+w}
\oplus S^t_{2,i+w} \cdot S^t_{3,i+w} \oplus S^t_{3,i+w} \cdot S^t_{2,i+2w}\notag\\
\oplus\; &S^t_{2,i+2w} \cdot S^t_{3,i+2w} \oplus S^t_{3,i+2w} \cdot S^t_{2,i+3w}
\oplus S^t_{2,i+3w} \cdot S^t_{3,i+3w} \oplus S^t_{3,i+3w} \cdot S^t_{2,i}.\label{eq:8circle}
\end{align}
It may be observed that the products occurring in the equation above involve 8 terms in a ring. The weight of this expression can be computed by brute force, and is equal to $3$.

@@

In the end, for \cipher{MORUS1280}, since the trail fragment $\gamma^t_i$ is used four times, the phenomenon we just described adds a contribution of $4 \cdot 3 = 12$ to the overall weight of the full trail. This results in a total weight of $4 \cdot 16 + 12 = 76$ (recall that the weight of the trail on \cipher{MiniMORUS1280} is 16). We have confirmed this bias by explicitly computing the full trail equation in \Cref{sec:traileq}, and evaluating its exact weight like we did for \cipher{MiniMORUS} in \Cref{sec:minibias}. That is, since the equation is quadratic, we may view it as a graph, which we split into connected components; we then compute the bias of each connected component separately by brute force, and then add up the weights of all components per the piling-up lemma. In the end, the full trail equation given in \Cref{sec:traileq} yields a weight of 76 for the full trail on \cipher{MORUS1280}.

In the case of \cipher{MORUS640}, collisions between rotation constants further complicate the analysis. Specifically, when using trail fragment $\beta^t_i$, the term $S^t_{2,i} \cdot S^t_{3,i}$ occurs. As explained previously, a partial collision with the term $S^t_{2,i} \cdot S^t_{3,i-w}$ from trail fragment $\gamma^t_i$ results in \Cref{eq:8circle}. However trail fragment $\alpha^t_{i+d}$ is once used in the course of the full trail with an offset of $d = b_1+b_4-b_0-b_2$ (relative to $\gamma^t_i$), which in the case \cipher{MORUS640} is equal to $31+13-5-7 = 0 \;\text{mod}\; 32$. This creates another term $S^t_{2,i} \cdot S^t_{3,i}$, which ultimately destroys one of the four occurrences of \Cref{eq:8circle}. In the end, when computing the full trail equation on \cipher{MORUS640}, we get that the weight of the trail is 73 (cf. \Cref{sec:traileq}).
